Binary Image Transition State Search Method

### 2022-03-30 manuscript.tex
started updating methods

\section{Method}
The method begins by first initialising the states, $\bm{x}_1$ and $\bm{x}_2$, in the basins of attraction of different local minima.
These can be set to the minima, but this is not a necessary requirement.
The BITSS method then leads these states to meet at the lowest point on the ridge between the basins of attraction - the transition state, as shown in \cref{fig:toy2d}a.
This is done by minimising the energy of the pair while constraining their separation and energy difference.
The separation constraint $d_i$ is incrementally reduced, such that at iteration $i$,
\begin{equation}\label{eq:diteration}
  d_i = (1 - f) d_{i-1},
\end{equation}
with $d_0$ the separation between the initial states.
A factor of $f = 0.5$ is succesful for most applications, but this can be made smaller to ensure that the states do not slide off the ridge.
The secondary constraint sets the energies of the two states equal to ensure that neither state is pulled over the ridge.

The two constraints are implemented using energy penalty terms, which, if the constraints are not met, result in driving forces on the two states, such as in \cref{fig:toy2d}b.
The total energy of the pair of states is therefore given by
\begin{multline}\label{eq:bitss}
  E_\text{BITSS}(\bm{x}_1, \bm{x}_2) = E_1 + E_2
    + \kappa_e \left( E_1 - E_2 \right) ^2 \\
    + \kappa_d \left( \mathrm{d}(\bm{x}_1, \bm{x}_2) - d_i \right) ^2,
\end{multline}
where $E_1$ and $E_2$ are the single state energies, $\kappa_e$ and $\kappa_d$ parametrise the strengths of the energy and distance constraints, and $\mathrm{d}(\bm{x}_1, \bm{x}_2)$ is the measure of distance between the two states.
In this work the L-BFGS algorithm is chosen to minimise the energy, owing to its fast convergence and low memory requirement for large numbers of degrees of freedom \cite{Liu1989}.
However, any other minimisation method can be used instead.

@@

To ensure that the transition state is located successfully, the constraint strengths $\kappa_d$ and $\kappa_e$ are updated as the algorithm proceeds using information from the system.
They are initially calculated at the start of each minimisation, and regularly recalculated throughout (once per 100 iterations is used in this work).
They are set using the following equations such that each term in the gradient of \cref{eq:bitss} is of a similar size,
\begin{gather}
  \kappa_e = \frac {\alpha} {2 E_B},
  \label{eq:ke}
  \\
  \kappa_d = \max \left(
    \frac {\sqrt{|\bm{\nabla} E_1|^2 + |\bm{\nabla} E_2|^2}} {2\sqrt{2} \beta d_i} \; , \;
    \frac{E_B}{\beta d_i^2} \right),
  \label{eq:kd}
\end{gather}
where $\bm{\nabla} E_1$ and $\bm{\nabla} E_2$ are the gradients of the energies of the two states, and $\alpha$ and $\beta$ are parameters with recommended values of $\alpha = 10$ and $\beta = 0.1$ (see supplementary information \footnote{See Supplemental Material at [URL] for derivations of the constraint coefficients, extra details for the example systems, results for the speed test against DNEB, and a comparison of the discontinuous energy landscape with the standard Lennard-Jones cluster}).
Here, $E_B$ is an estimation for the current energy barrier; evaluated using the difference between the highest energy along a linear interpolation between the two states and the average energy of the two states.

In summary, the method involves iteratively performing the following three steps:
\begin{enumerate}[noitemsep,nolistsep]
  \item Reduce the target separation using \cref{eq:diteration}.
  \item Compute the constraint coefficients using \cref{eq:ke,eq:kd}.
  \item Minimise the potential, \cref{eq:bitss}, and recompute the coefficients regularly throughout the minimisation.
\end{enumerate}
This process is completed once a convergence criterion is reached that can be based upon either the separation between the states, or the size of gradient at the midpoint between the two states.

@@

\section{Results}
\subsection{Speed Comparison}
\begin{figure*}[tb]
  \includegraphics{speedtest.pdf}
  \caption{\label{fig:speedtest}
    Comparisons between BITSS, DNEB, and the string method in the convergence to the transition state as a function of the number of calculations of the energy gradient.
    Three test systems are used: (a) a Lennard-Jones seven-particle cluster, (b) cylindrical shell buckling, and (c) wetting of a chemically-striped surface.
    The configurations shown correspond to the two minimum energy states and the transition state, marked by an asterisk.
    The string method and DNEB are repeated with a differing number of images, as listed in the legend in (c).
  }
\end{figure*}
As a result of the BITSS method using only two states it has the potential to be significantly faster than chain-of-states approaches which typically use tens of states.
This is tested here by measuring the speed of convergence to the transition state for BITSS in comparison with the string method and DNEB for several representative examples.

@@

The results for this test, shown in \cref{fig:speedtest}, demonstrate two main advantages for the BITSS method.
Firstly, the slope illustrates that the rate of convergence is indeed much greater for the BITSS method.
It is matched only by the initial rate of the string method using less than ten images, however these soon slow and are relatively inaccurate.
Secondly, the BITSS method can more accurately approximate the transition state by orders of magnitude, particularly for the wetting example where the string method does not significantly improve upon the initial pathway estimate.
Therefore, unlike the string method, BITSS does not require a second approach to accurately locate the transition state.
Similar results are also found when comparing against against DNEB, which can be seen in the supplementary information \cite{Note1}.

@@

\subsection{Locally flat landscapes}
\begin{figure}[tb]
  \includegraphics{flatdiscontinuous.pdf}
  \caption{\label{fig:flatdiscontinuous}
    (a) Energy profile of the BITSS pathway on a 2D potential with flat regions.
        Blue and red dots denote the minima and transition state, respectively.
        Points of interest are labelled by i--iii (see text).
        Top inset: A zoomed in view around the transition state.
        Bottom inset: The pathway taken, with the edges of the flat regions marked by dashed lines.
    (b) Disconnectivity graph of the energy landscape for a seven-particle cluster with a discontinuous hard-core pair potential (orange), and a standard Lennard-Jones potnetial (gray).
        The two graphs are offset for visibility.
    (c) Plot of the discontinuous potential.
        The repulsive section of the standard Lennard-Jones potential is also shown in grey.
  }
\end{figure}

@@

\subsection{Undefined gradients}
Finally, we investigate the application of BITSS to systems with undefined gradients, such as when the landscape is discontinuous.
In this case a couple of alterations to the method must be made.
First, the gradients in \cref{eq:kd} are ill-defined, so $\kappa_d$ can be calculated using just the second term in the equation.
Next we must use a minimiser that does not require a differentiable optimisation function; for this we use simulated annealing \cite{Kirkpatrick1983}.
This has a chance of randomly jumping one state over the dividing barrier, but we can reduce this probability by limiting the initial temparature and maximum random displacement. We typically employ $T_0 = E_B / 10$, and $d_\text{max} = d(\bm{x}_1, \bm{x}_2) / 100$.
This has been tested using a Lennard-Jones 7-particle cluster with hard cores, resulting in a discontinuous landscape.
Using the gradient-free approach, BITSS is able to successfully find the transition states, allowing us to plot the disconnectivity graph of the system, shown in \cref{fig:flatdiscontinuous}b.
A comparison with the results for the standard Lennard-Jones cluster is given in the supplementary information \cite{Note1}.
It is worth noting, however, that a gradient-based approach is significantly more efficient, and so should be preferred if possible.

@@

>>>>

\section{Results and discussion}
\subsection{BITSS method in a nutshell}
The method begins by first initialising the states, $\bm{x}_1$ and $\bm{x}_2$, in the basins of attraction of different local minima.
These can be set to the minima, but this is not a necessary requirement.
The BITSS method then leads these states to meet at the lowest point on the ridge between the basins of attraction - the transition state, as shown in \cref{fig:toy2d}a.
This is done by minimising the energy of the pair while constraining their separation and energy difference.
The separation constraint $d_i$ is incrementally reduced, such that at iteration $i$,
\begin{equation}\label{eq:diteration}
  d_i = (1 - f) d_{i-1},
\end{equation}
with $d_0$ the separation between the initial states.
A factor of $f = 0.5$ is succesful for most applications, but this can be made smaller to ensure that the states do not slide off the ridge.
The secondary constraint sets the energies of the two states equal to ensure that neither state is pulled over the ridge.

The two constraints are implemented using energy penalty terms, which result in driving forces on the two states if the constraints are not met, such as in \cref{fig:toy2d}b.
The total energy of the pair of states is therefore given by
\begin{multline}\label{eq:bitss}
  E_\text{BITSS}(\bm{x}_1, \bm{x}_2) = E_1 + E_2
    + \kappa_e \left( E_1 - E_2 \right) ^2 \\
    + \kappa_d \left( \mathrm{d}(\bm{x}_1, \bm{x}_2) - d_i \right) ^2,
\end{multline}
where $E_1$ and $E_2$ are the single state energies, $\kappa_e$ and $\kappa_d$ parametrise the strengths of the energy and distance constraints, and $\mathrm{d}(\bm{x}_1, \bm{x}_2)$ is the measure of distance between the two states.
In this work the L-BFGS algorithm is chosen to minimise the energy, owing to its fast convergence and low memory requirement for large numbers of degrees of freedom \cite{Liu1989}.
However, any other minimisation method can be used instead.

In summary, the method involves iteratively performing the following three steps:
\begin{enumerate}[noitemsep,nolistsep]
  \item Reduce the target separation using \cref{eq:diteration}.
  \item Compute the constraint coefficients using \cref{eq:ke,eq:kd}.
  \item Minimise the potential given by \cref{eq:bitss}, and recompute the coefficients at regular intervals.
\end{enumerate}
This process is completed once a suitable convergence criterion is reached.
This can either be based upon the separation between the states, or the size of gradient at the midpoint between the states.

@@

@@

\subsection{Speed Comparison}
\begin{figure*}[tb]
  \includegraphics{speedtest.pdf}
  \caption{\label{fig:speedtest}
    Comparisons between BITSS, DNEB, and the string method in the convergence to the transition state as a function of the number of calculations of the energy gradient.
    Three test systems are used: (a) a Lennard-Jones seven-particle cluster, (b) cylindrical shell buckling, and (c) wetting of a chemically-striped surface.
    The configurations shown correspond to the two minimum energy states and the transition state, marked by an asterisk.
    The string method and DNEB are repeated with a differing number of images, as listed in the legend in (c).
  }
\end{figure*}
As a result of the BITSS method using only two states it has the potential to be significantly faster than chain-of-states approaches which typically use tens of states.
This is tested here by measuring the speed of convergence to the transition state for BITSS in comparison with the string method and DNEB for several representative examples.

@@

The results for this test, shown in \cref{fig:speedtest}, demonstrate two main advantages for the BITSS method.
Firstly, the slope illustrates that the rate of convergence is indeed much greater for the BITSS method.
It is matched only by the initial rate of the string method using less than ten images, however these soon slow and are relatively inaccurate.
Secondly, the BITSS method can more accurately approximate the transition state by orders of magnitude, particularly for the wetting example where the string method does not significantly improve upon the initial pathway estimate.
Therefore, unlike the string method, BITSS does not require a second approach to accurately locate the transition state.

@@

\subsection{Complex landscapes}
\begin{figure}[tb]
  \includegraphics{flatdiscontinuous.pdf}
  \caption{\label{fig:flatdiscontinuous}
    (a) Energy profile of the BITSS pathway on a 2D potential with flat regions.
        Blue and red dots denote the minima and transition state, respectively.
        Points of interest are labelled by i--iii (see text).
        Top inset: A zoomed in view around the transition state.
        Bottom inset: The pathway taken, with the edges of the flat regions marked by dashed lines.
    (b) Disconnectivity graph of the energy landscape for a seven-particle cluster with a discontinuous hard-core pair potential (orange), and a standard Lennard-Jones potnetial (gray).
        The two graphs are offset for visibility.
    (c) Plot of the discontinuous potential.
        The repulsive section of the standard Lennard-Jones potential is also shown in grey.
  }
\end{figure}

@@

Finally, we investigate the application of BITSS to systems with undefined gradients, such as when the landscape is discontinuous.
In this case a couple of alterations to the method must be made.
First, the gradients in \cref{eq:kd} are ill-defined, so $\kappa_d$ can be calculated using just the second term in the equation.
Next we must use a minimiser that does not require a differentiable optimisation function; for this we use simulated annealing \cite{Kirkpatrick1983}.
This has a chance of randomly jumping one state over the dividing barrier, but we can reduce this probability by limiting the initial temparature and maximum random displacement. We typically employ $T_0 = E_B / 10$, and $d_\text{max} = d(\bm{x}_1, \bm{x}_2) / 100$.
This has been tested using a Lennard-Jones 7-particle cluster with hard cores, resulting in a discontinuous landscape.
Using the gradient-free approach, BITSS is able to successfully find the transition states, allowing us to plot the disconnectivity graph of the system, shown in \cref{fig:flatdiscontinuous}b.
A comparison with the results for the standard Lennard-Jones cluster is given in the supplementary information \cite{Note1}.
It is worth noting, however, that a gradient-based approach is significantly more efficient, and so should be preferred if possible.

@@

\section{Methods}
\subsection{BITSS: Target distance}

\subsection{BITSS: Total potential}

\subsection{BITSS: Constraint coefficients}
To ensure that the transition state is located successfully, the constraint strengths $\kappa_d$ and $\kappa_e$ are updated as the algorithm proceeds using information from the system.
They are initially calculated at the start of each minimisation, and regularly recalculated throughout (once per 100 iterations is used in this work).
They are set using the following equations such that each term in the gradient of \cref{eq:bitss} is of a similar size,
\begin{gather}
  \kappa_e = \frac {\alpha} {2 E_B},
  \label{eq:ke}
  \\
  \kappa_d = \max \left(
    \frac {\sqrt{|\bm{\nabla} E_1|^2 + |\bm{\nabla} E_2|^2}} {2\sqrt{2} \beta d_i} \; , \;
    \frac{E_B}{\beta d_i^2} \right),
  \label{eq:kd}
\end{gather}
where $\bm{\nabla} E_1$ and $\bm{\nabla} E_2$ are the gradients of the energies of the two states, and $\alpha$ and $\beta$ are parameters with recommended values of $\alpha = 10$ and $\beta = 0.1$ (see supplementary information \footnote{See Supplemental Material at [URL] for derivations of the constraint coefficients, and extra details for the example systems}).
Here, $E_B$ is an estimation for the current energy barrier; evaluated using the difference between the highest energy along a linear interpolation between the two states and the average energy of the two states.

\subsection{BITSS: Changes for undefined gradients}

\subsection{Adaptive discretisation test details}

\subsection{Particle cluster system}

\subsection{Cylindrical buckling system}

\subsection{Striped wetting system}
